---
title: "Analyzing Fannie Mae Loans Difference between 2007 and 2019"
author: Pin-Chen Chi
format: html
output: bookdown::html_document2
toc: false
editor: visual
---

```{=html}
<!This block assists in formatting the title, font size, chart caption, etc.– –>
<style type="text/css">
  .title {
    text-align: left;
}
body{ /* Normal  */
      font-size: 16px;
      margin: 2.5cm;
      top: 1.5cm;
      text-align: justify;
  }
  .caption {
    font-size: big;
    text-align: center;
    position: above;
}
</style>
```

```{css plotly-caption, echo = FALSE}
/*Include this block if you like the chart caption at the top instead of the bottom.*/
div.figure {
  display: table;
}
div.figure p {
  display: table-caption;
  caption-side: top;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

------------------------------------------------------------------------

In this report, I analyze the differences in loans owned by Fannie Mae
between 2007 and 2019, focusing on the changes in the default rates,
borrower credit scores, property occupancy status, debt-to-income
ratios, and interest rates. Based on the two datasets in 2007 and 2019,
I found that the default rate for loans was 9.02% in 2007, during the
mortgage crisis, while in 2019, the default rate had significantly
decreased to 0.2%, indicating a more stable period.\
\
Figure \@ref(fig:Figure1) shows each state’s percentage change of
default rate of loans relative to the national default rate in both
years. In 2007, some states had higher default rates exceeding the
national default rate(9.02%). For example, The default rate in Nevada
increased by 257.1% compared to the national default rate (9.02%) and
Arizona State increased by 174.5% compared to the national rate. This
suggests that these states were significantly more impacted by the
mortgage crisis in 2007; In 2019, default rates of west virginia state
and vermont state were significantly higher than national default
rate(0.2%).     

Next, I examine the average borrower credit scores in 2007 and 2019. The
average credit score in 2007 was 726.04, while in 2019 it had risen to
761.31(calculated in codes). Figure \@ref(fig:Figure2) shows the
distribution of credit scores for defaulted loans in 2007, with scores
concentrated between 650 and 700. In contrast, Figure \@ref(fig:Figure3)
shows that in 2019, credit scores were concentrated between 700 and 750,
indicating that borrowers had higher credit scores in 2019. This
suggests that in 2007, defaulted loans were more likely to come from
borrowers with relatively lower credit scores.

In Figure \@ref(fig:Figure4), I observe that the Debt-To-Income ratio
for loans in 2007 is concentrated around 40%, whereas in 2019 the ratios
are noticeably lower. This indicates that in 2007 a larger portion of
borrowers' incomes were allocated to debt payments. Since higher DTI
ratios are associated with greater financial risk, this may have
contributed to a higher likelihood of default during that period.\
\
In Figure \@ref(fig:Figure5), it shows that interest rates in 2007 were
generally higher than in 2019, suggesting that loans issued during the
mortgage crisis had higher borrowing costs. As a result, the financial
stress caused by these higher payments likely contributed to the
increased default rates in 2007. Next, I examine the default rates
across different property occupancy statuses. Finally, I examine Figure
\@ref(fig:Figure6). It suggests that in 2007, investor-mortgages
exhibited the highest default rate 14.3%, showing that
investor-mortgages were more likely to default. This can be explained
that investor-mortgages, which are not for primary residences purposes,
were more likely to default, especially in the loan crisis year.
Besides, the average interest rate of investor-mortgages is
6.89%(calculated in codes), which is higher than the interest of two
others(principal-mortgage: 6.45%; second-mortgage: 6.46%). This supports
the above idea that higher interest rates are correlated with higher
default risk. In 2019, default rates across all property types were
significantly lower. Investor-mortgages had a default rate of 0.2%,
principal-mortgages 0.2%, and second-mortgages 0%, reflecting the
overall low default rate in 2019.

Overall, above data visualizations reveal the contrast between loan
performance in 2007 and 2019: In 2007, the default rate was much higher,
driven by factors like elevated interest rates, higher Debt-to-Income
ratios, and lower borrower credit scores. Compared to 2019, the economic
conditions in 2007 during the mortgage crisis caused considerable
financial stress for a large portion of borrowers, making it more
challenging for them to meet their loan obligations.

(Word count: 609)

<br>

### Figure Appendix {.unnumbered}

```{r, include = FALSE}
library(readr)
library(tidyverse)
library(readxl)
library(scales)
library(zoo)
library(patchwork) # package for combining charts
library(usmap)  # package for plotting maps 
library(plotly) #package for simple interactive charts
library(ggiraph)

```

```{r the label of your chart, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="the caption of your chart"}
# This is the code block for producing charts. 
# You can include as many of these blocks as needed. 
# By adding "echo = FALSE," only the chart and not the code will be included in the output HTML file.
# To modify the size of your chart, adjust the values of fig.width and fig.height.
# Use fig.cap to specify the caption for your chart.
# You can insert a blank line between charts by adding "<br>" below the block.
```

<br>

```{r, include = FALSE}
data_2007 = read_rds('data_sample_2007Q4.rds')
data_2007$YEAR <- 2007
data_2007
```

```{r, include = FALSE}
data_2019 = read_rds('data_sample_2019Q4.rds')
data_2019$YEAR <- 2019
data_2019
```

```{r, include = FALSE}
# Only defaluted loans: DEFAULT_FLAG = 1 
data_2007_default <- data_2007 %>%
  filter(DEFAULT_FLAG == 1)

data_2007_default
# default rate of 2007 Q4 is 4508/50000 = 9.02% -> 0.0902

data_2019_default <- data_2019 %>%
  filter(DEFAULT_FLAG == 1)

data_2019_default
# default rate of 2019 Q4 is 100/50000 = 0.2% -> 0.002
```

```{r, include = FALSE}

# default rate of 2007 Q4 is 4508/50000 = 9.02% -> 0.0902
# default rate of 2019 Q4 is 100/50000 = 0.2% -> 0.002

default_state_2007 <- data_2007 %>%
  group_by(STATE) %>%
  summarise(defaultRate_state = mean(DEFAULT_FLAG, na.rm = TRUE)) %>%
  mutate(defaultIndex_state = ((defaultRate_state - 0.0902) / 0.0902) * 100) #percentage

default_state_2007 <- default_state_2007 %>%
  rename(state = STATE)
default_state_2007

default_state_2019 <- data_2019 %>%
  group_by(STATE) %>%
  summarise(defaultRate_state = mean(DEFAULT_FLAG, na.rm = TRUE)) %>%
  mutate(defaultIndex_state = ((defaultRate_state - 0.002) / 0.002) * 100) #percentage

default_state_2019 <- default_state_2019 %>%
  rename(state = STATE)
default_state_2019

```

```{r Figure1, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Increase or Decrease of Default Rates by states relative to national in 2007(left) vs. 2019(right)"}
map_2007 <- plot_usmap(regions = "state", data = default_state_2007, 
                          values = "defaultIndex_state", color = "white") + 
  scale_fill_gradient(name="Relative to National Default Rate(%)", 
                    limits = c(min(default_state_2007$defaultIndex_state), 
                               max(default_state_2007$defaultIndex_state)),
                      low = "lightblue", high = "blue")+
  aes(text = paste("State:", state, "<br>",
                   "Relative to National Default Rate:", round(defaultIndex_state, 1), "%"))


map_2019 <- plot_usmap(regions = "state", data = default_state_2019, 
                           values = "defaultIndex_state", color = "white") + 
  scale_fill_gradient(name="Relative to National Default Rate(%)", 
                 limits = c(min(default_state_2019$defaultIndex_state), max(default_state_2019$defaultIndex_state)),
                      low = "lightblue", high = "blue")+
  aes(text = paste("State:", state, "<br>",
                   "Relative to National Default Rate:", round(defaultIndex_state, 1), "%"))

#both maps
subplot(ggplotly(map_2007), ggplotly(map_2019))
```

```{r, include = FALSE}
# averag credit score
avg_CS_2007 <- mean(data_2007$CSCORE_C, na.rm = TRUE)
avg_CS_2007 #726.0368
avg_CS_2019 <- mean(data_2019$CSCORE_C, na.rm = TRUE)
avg_CS_2019 #761.3086
```

<br>

<br>

```{r Figure2, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Borrower Credit Score of Defaluted loans in 2007"}
data_2007_default %>%
  ggplot(aes(x=CSCORE_B)) + 
  geom_histogram(breaks=seq(500, 900, 50), fill="lightblue", color="white") + 
  scale_x_continuous(breaks = seq(500, 900, 50))+
  labs(title=" ",
       x="Borrower Credit Score", y="Frequency") +
  theme_classic() +
  theme(axis.title.x = element_text(margin = margin(t = 10)), 
        plot.title = element_text(size=12),
        panel.grid.major.y = element_line(color="grey95"))

#In 2007 Q4, Borrower Credit Score is concentrated between 650 and 700

```

<br>

<br>

```{r Figure3, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Borrower Credit Score of Defaluted loans in 2019 Q4"}
data_2019_default %>%
  ggplot(aes(x=CSCORE_B)) + 
  geom_histogram(breaks=seq(500, 900, 50), fill="royalblue", color="white") + 
  scale_x_continuous(breaks = seq(500, 900, 50))+
  labs(x="Borrower Credit Score", y="Frequency") +
  theme_classic() +
  theme(axis.title.x = element_text(margin = margin(t = 10)), 
        plot.title = element_text(size=12),
        panel.grid.major.y = element_line(color="grey95")) 

# In 2019 Q4, Borrower Credit Score is concentrated between 700 and 750

```

```{r, include = FALSE}

#data: combine 2007 Q4 data & 2019 Q4 data

data_2007 = read_rds('data_sample_2007Q4.rds')
data_2007$YEAR <- 2007

data_2019 = read_rds('data_sample_2019Q4.rds')
data_2019$YEAR <- 2019

data <- bind_rows(data_2007, data_2019)
```

```{r, include = FALSE}
avg_DTI_2007 <- mean(data_2007$DTI, na.rm = TRUE)
avg_DTI_2007 #38.56981
avg_DTI_2019 <- mean(data_2019$DTI, na.rm = TRUE)
avg_DTI_2019 #34.58998
```

<br>

<br>

```{r Figure4, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Analyzing Debt-To-Income Ratio between 2007 vs. 2019"}

set.seed(201)
data %>%   
  mutate(`Year` = ifelse(YEAR == 2007, "2007", ifelse(YEAR == 2019, "2019"))) %>% 
  #mutate(`Year` = factor(`Year`, levels = c("2007", "2019"))) %>%  # Ensure proper ordering
  
  ggplot(aes(x = `Year`, y = DTI, col = `Year`)) +  # Use 'Year' for color mapping

  # Two different colors (as the 'Year' column will control the color)
  geom_jitter(width = 0.1, height = 0, stroke = NA, alpha = 0.5, size = 2) +  
  geom_boxplot(alpha = 0.1, outlier.color = NA) +  # Boxplot
  
  labs(title = "",
       x = "Year", y = "Debt-To-Income Ratio(%)") +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.title.x = element_text(margin = margin(t = 20)),  # Adjust the x-axis title spacing
        panel.grid.major.y = element_line(color = "grey95")) +  # Light grey grid lines
  
  scale_color_brewer(palette = "Set2") +  # Set color palette
  scale_y_continuous()  # Adjust y-axis scale
```

<br>

<br>

```{r Figure5, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Comparing Interest Rate on Loans in 2007 vs. 2019"}

set.seed(201)
data %>%   
  mutate(`Year` = ifelse(YEAR == 2007, "2007", ifelse(YEAR == 2019, "2019"))) %>% 
  
  ggplot(aes(x = `Year`, y = ORIG_RATE, col = `Year`)) +  

  geom_jitter(width = 0.1, height = 0, stroke = NA, alpha = 0.5, size = 2) +  
  geom_boxplot(alpha = 0.1, outlier.color = NA) +  # Boxplot
  
  labs(title = "",
       x = "Year", y = "Original Interest Rate") +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.title.x = element_text(margin = margin(t = 20)),  # Adjust the x-axis title spacing
        panel.grid.major.y = element_line(color = "grey95")) +  # Light grey grid lines
  
  scale_color_brewer(palette = "Set1") +  # Set color palette
  scale_y_continuous() 
```

```{r include=FALSE}
OLTV_2007 <- data_2007 %>%
  filter(DEFAULT_FLAG == 1) #%>%

avg_OLTV_2007 <- mean(OLTV_2007$OLTV, na.rm = TRUE)
avg_OLTV_2007

OLTV_2019 <- data_2019 %>%
  filter(DEFAULT_FLAG == 1) #%>%

avg_OLTV_2019 <- mean(OLTV_2019$OLTV, na.rm = TRUE)
avg_OLTV_2019
```

```{r include=FALSE}
# prepare data

OCC_STAT_All <- data %>%
  group_by(YEAR, OCC_STAT) %>%
  summarise(default_rate = mean(DEFAULT_FLAG, na.rm = TRUE), .groups = 'drop')

OCC_STAT_All <- OCC_STAT_All %>%
  mutate(OCC_STAT_2 = case_when(
    OCC_STAT == "I" ~ "Investor",
    OCC_STAT == "P" ~ "Principal",
    OCC_STAT == "S" ~ "Second",
    OCC_STAT == "U" ~ "Unknown",
  ))

# View the summarized data
OCC_STAT_All
```

<br>

<br>

```{r Figure6, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Comparing Default Rate of property occupancy status in 2007 vs. 2019"}

OCC_STAT_All %>%
  ggplot(aes(x=as.factor(YEAR), y=default_rate, fill = OCC_STAT_2)) +
  geom_col(width=0.7, position="dodge")+
  labs(y = "Default Rate",
       x = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.line.y = element_line(),
        panel.grid = element_blank(),
        panel.grid.major.y = element_line(color = 'grey90'),
        axis.text = element_text(size=11)) + 
  geom_text(aes(y = default_rate, label = paste0(round(default_rate * 100, 1), "%")), 
            position = position_dodge(0.7)) +
  scale_fill_brewer(name = '', palette = 'Set2')
```

```{r include=FALSE}
OCC_avg_interestRate <- data %>%
  group_by(YEAR, OCC_STAT) %>%
  summarise(avg_interestRate = mean(ORIG_RATE, na.rm = TRUE), .groups = 'drop')

OCC_avg_interestRate
```
